\documentclass[11pt]{article}

\usepackage{xltxtra}
\setmainfont[Mapping=tex-text]{Linux Libertine O}
\usepackage{polyglossia}
\setmainlanguage[variant=british]{english}

\usepackage{ledmac}
\usepackage{ledpar}
\usepackage{csquotes}
\usepackage{setspace}

\noendnotes

\onehalfspace

\author{Christopher Guy Yocum}
\title{\emph{Finn\.{s}ruith F\'{i}thail}: Text A (TCD H. 3. 18)}

\footparagraph{A}

\begin{document}

\maketitle

{\Large \textbf{THIS IS A FIRST DRAFT TRANSLATION} as such it should be treated with extreme caution.  There will be errors and omissions of all forms.  Some of these errors may be quite amusing.  Be that as it may, the translation will be updated from time to time with more and better translations and notes.}

The \emph{Finn\.{s}ruith F\'{i}thail} is the first and oldest text to contain any traces of the early Irish judge F\'{i}thal.  Moreover, it is an early Irish legal text which has remained without a translation or a critical edition.  This draft translation moves forward knowledge both of F\'{i}thal in particular but early Irish law more generally and allows scholars to delve confidently into the particulars of early Irish law.

\emph{Finn\.{s}ruith F\'{i}thail} has a diplomatic transcription by D.A. Binchy in his magisterial \emph{Corpus Iuris Hibernici} where it appears in two versions: A (CIH786.25--789.17) and B (CIH 2131.1--2143.40).  The translation offered here is based exclusively on Binchy's transcription of the A text.

\section{Previous Scholarship}

[insert stuff here]

\section{Editorial Methodology}



\section{Dating}

\newpage

\begin{pages}
  \begin{Leftside}
    \beginnumbering
    
    \pstart
    A mo sruith .i. ol socht. Co ber br\emph{eith} .i. cionnus berat b\emph{r}eatha.  b\emph{er}o nip sl\emph{an} .i. ar fithal.  nach saithi .i. diub\emph{air}t na mboeth.  ni hiadhat \edtext{\emph{ratha}}{\Afootnote{.r. = ratha}} for diuburta .i. ni is mo na coruice leth; dofuasl\emph{aic}e fir cach doch\emph{ur} .i. cuir na memar timcheall na ceann. 
    \pend

    \pstart
    Ni \edtext{\emph{saire}}{\Afootnote{saire = saighre}} .i. l.at no .xl.at [\hspace{2mm}] na rann .i. rann de resin \edtext{re-sin,}{\Afootnote{repeat?}} no ini rannait na braith\emph{r}i eturr\emph{a} co ceann .u. mbliad\emph{na}: feri\emph{ann} dib\emph{aid} toebfine.  Na rathus .i. i ndilsi in seoitt. na hairgsi .i. duine truaigh.  na haititiu .i. annfesa na codnach.  fo\emph{r} \edtext{\emph{dichell}}{\Afootnote{.d. = dichell}} .i. s. sl\emph{an} do choinn .s. essl\emph{ain}. imteca\emph{r}tha\emph{r} cac\emph{c} leithchil \edtext{\emph{cloen}}{\Afootnote{.c. = cloen}} .i. em-teclamtha\emph{r} ca\emph{ch} lethcloen | i cutuma ndligid.  Arnacon dearba\emph{r}a nech ar.ii. \edtext{.i. Arnar .i.}{\Afootnote{\emph{dittogr.?}}} Arna raibh diub\emph{air}t nech dibh oc ailaile.
    \pend
    
    \pstart
    \textbf{IS} amnech .i. is aml\emph{aid}-seo[\hspace{2mm}]for nadm\emph{an} .i. f\emph{or} treb\emph{air}e[\hspace{2mm}]cen easbuidh .i. i mbec.  cen tessb\emph{aid} .i. i mor[\hspace{2mm}]nibi cor .i. itir dhiss.  nibi naidm .i. treb\emph{aire} echtrann.  m\emph{ain}e comlinat fol\emph{ad} .i. lanloighi. r\'{i} cach a co\emph{m}ann .i. is righ cach f\emph{or}in \edtext{bioss}{\Afootnote{?}} aici, ɫ is innrigthigh cach im ath\emph{gabail} do gab\emph{ail} i cae cen mann iar nap\emph{ad} ⁊ iar trosca.  saer cach a nadma .i. is saml\emph{aid}-seo is saer do cach cuiti \edtext{treb\emph{air}i}{\Afootnote{followed by erasure (about 5 letters)}} d'fas\emph{tad}(?)
    \pend
    \endnumbering
  \end{Leftside}

  \begin{Rightside}
    \beginnumbering
    
    \pstart
    \enquote{Oh my elder} .i. says Socht.  How should/shall I judge .i. how does/shall he bear judgements.  I should/shall bear [judgement] such that it may not be exempt .i. says F\'{i}thal. so that there is no throng .i. the fraud of the legally incompetent.  He does not end [a transaction on] sureties on account of fraud .i. it is not which is greater than half disgrace; men dissolve each disadvantageous contract .i. the members put a boundary [on] the heads.   
    \pend
    
    \pstart
    It is not the exemption .i. 40th or 50th of the parts(?). of the parts(?) .i. a part from before the previous(?), or those(?) brothers divide among them up to the end of fifth year: land legacy of the side-kingroup. The suretyship .i. in the securityship of a unit of value.  of the difficulty .i. the miserable man (person?). of the acknowledging of property rights .i. ignorance of the sensible adult. for negligence .i. s. exempt in return for(?) .s. out of [the] whole. every half-fault of iniquity .i. each half-iniquity gathers | in equal rights. that he should not defraud anyone [\ldots] that fraud was not anyone from them at the other. 
    \pend
  
    \pstart
    It is indeed anyone .i. Thus on a surety .i. a guarantee without deficiency .i. in a small without absences .i. in large it is not a surety .i. between two.  it is not a pledge .i. a stranger's pledge.  of cattle, completing means of liability .i. full lying down(?).  each king his cattle-pound [\ldots] at it(?), or it is each king's house to take concerning distrait in visiting without an ingot(?) after proclaiming and fasting.  each free-client his surety .i. thus it is it a free-client to establish(?) for each portion of a pledge.

    \pend
  \endnumbering
  \end{Rightside}
  \Pages
 
  \begin{Leftside}
    \beginnumbering
   \pstart 
    Manip dlig\emph{ed} dociallath\emph{ar} \edtext{folu}{\Afootnote{folad(?) DIL D2 230.065}} fir .i. manib e dlig\emph{ed} dogne se, adbul-dich\emph{ell} in fir aili do reir firine ima folu, ni is \edtext{mo}{\Afootnote{niis mo \emph{repeated (dittogr.) cancelled by underlining}}} na cuiti trebairi do fasta.  Ar is ferr fir firsiu .i. fol\emph{ad} .l. na f\emph{er}tsechus n\emph{ir}t.  is ferr folad .i. is ferr fia d\'{i}a fol\emph{ad} comtoiridhne na ast\emph{ad} fesa ⁊ t\emph{r}eb\emph{air}i.  \edtext{\emph{f\emph{or} fonaidm}}{\Afootnote{\emph{sic}, fonaim}} .i. t\emph{r}eb\emph{air}i cen fol\emph{ad}.  aititiud .i. na cend ar aird.  degnadmaim .i. na t\emph{re}b\emph{air}i cen fol\emph{ad}.  \edtext{F\emph{or}congair}{\Afootnote{\emph{followed by erasure (two or three letters)}}} nad aurgair .i. is f\emph{or}cong\emph{r}a d\'{o} man rourgair oc den\emph{am} in cuir in tan na tarthaidt\emph{er} ant\'{i} dogn\'{i}.  ɫ na b\'{i} occo; ɫ ma fla\emph{ith} ɫ ecl\emph{ais} leicit cen a taith\emph{mech}.  Eirne contuaisi .i. ic\emph{aid} ma fine bess i nurrtuasacht.  Fuidhri fl\emph{ath}a .i. særfuigir ⁊ doerfuidhri.  ni astaith\emph{er} sæithiu .i. .s. sla\emph{n} t\emph{ar} cen\emph{n} .s. essl\emph{ain}.  sech a cundu .x. (?) .i. sech chomairli a codnach. 
    \pend
  
    \pstart
    \textbf{C}o b\emph{er} br\emph{eith} im indile. Ma dethb\emph{ir}us fir fin\emph{n}e, a caithem d\'{o} 'aonur co miss iar samai\emph{n}, ⁊ \edtext{sacar\emph{ad}}{\Afootnote{or scar\emph{ait}?}} iar s\emph{in}.  Mad indethb\emph{ir}us, a caithem do-som a oenur co samuin ⁊ comcaithim doibh a ndis co mis iar sin.  Ma dethb\emph{ir}us fir anfine, a caith\emph{em} dhoibh a ndis co mis iar samhain.  Ma \edtext{inndeich.x.beirus}{\Afootnote{sic (dittogr.)}} fir \edtext{\emph{anfine}}{\Afootnote{.a. = anfine}}, a chaith\emph{em} do a aenar co sam\emph{ain}, scaru o sam\emph{ain} amach.  Dia trebho ⁊ do chaith\emph{em} a \.{f}e\'{o}ir tuco in fer an\emph{n} so | .i. trian cach neich infasas \edtext{⁊rl-}{\Afootnote{et reliqua}}[\hspace{2mm}]a bun foraind .i. amail a barr.  Ar is cach dia feor .i. is i\emph{n}t\'{i} doratt for iasacht \edtext{dia}{\Afootnote{a subscript, added later; an leg. di ara thir?}} arathir o samoin amach.  Co b\emph{er} br\emph{eith} im techta .i. tir dib\emph{aid} so.  \edtext{Comairme}{\Afootnote{\emph{really belongs to preceding brocard, cf. 2135.37}}} .i. cutrumas airme.  Comecor i tir dib\emph{aid}.  Comairimh a fer\emph{ann} ath\emph{ar} ⁊ senath\emph{ar}.  Ma ruc nech ima\emph{r}craidh a tir dib\emph{aid}, no d\'{a} mil as ecuttruma a\'{i}s a nag\emph{aid} ainmil, is fiach duninicaiche cin\emph{n}te fair.
    \pend
  \endnumbering
  \end{Leftside}

\begin{Rightside}
    \beginnumbering
    \pstart
    If it is not a legal right which neglects a man's substance .i. if it is not a legal right which it may do, strongly neglecting the other man on the authority of a just [ruling?] concerning his substance, it is not greater to establish a portion of a security.  For strong men are better .i. full material of the [\ldots] strength. material is better .i. it is better material of the legally fixed shared lead water of knowing and pledge(s?) in the presence of God.  on a pledging .i. the securities without material. acknowledging .i. the heads at it.  in return for a pledge .i. the securities without material.  he commands that not forbid .i. it is a command to him to do a good verbal contract which he forbade when the aforementioned does the [\ldots].  or which he does not have; or if a greater lord or church leaves without the annulling.  he bestows [and] listens .i. if a kin pay which maybe kept silent.  A tenant of a lord .i. free-tenant and unfree-tenant. He is not establishing more hardship .i. [\ldots] exempt over an end(?) [\ldots] out of the whole.  without his friendship .x.(?) .i. without advice of his competent adult 
    \pend
    
    \pstart
    how should/shall I judge concerning property. if [\ldots] a true kin group, when spending to him alone up to a month after Samain, and separating after that. if it may be [\ldots] his using up to him his alone to Samain, together using up to them their standing(?) up to a month after that.  If [\ldots] true uncorporate kin group, his using up to them their standing(?) up to a month after Samhain.  If a true uncorporate kin, his using up to him his alone up to Samain, spreading out from Samain out.  If by means of a cow and to use up his grass [\ldots] the man here.  a third part of each one increases food-rent [and the rest].[\hspace{2mm}] his base on us .i. like his end. For it is each if grass(?) .i. it is he who has given on a loan if [\ldots] Samhain out.  how shall/should I judge concerning legal rightness .i. land of inheritance.  [\ldots] counting.  [\ldots] in land of inheritance.  proportional stock of their father and grandfather's land. if anyone will carry excess of their inheritance land, or two animals of of [\ldots]-heavy of years (their face driving animal?) it is a definite obligation of trespass by person on it. 

    \pend
  \endnumbering
  \end{Rightside}

  \Pages
 
  \begin{Leftside}
    \beginnumbering
    \pstart
    Co b\emph{er} br\emph{eith} a tairrsce .i. \edtext{caei}{\Afootnote{\emph{a subscript added later}}} tairis.  Gab\emph{ae}(?) ma 'troiris .i. ath\emph{gabail} bruighrechta[\hspace{2mm}]cor is c\'{o}ir .i. iar ndenumh a chinaidh.  Co cinaidh iar foc\emph{ru} .i. is iar foc\emph{ra} cinaidh ataitt na meich no in \edtext{duinicaichi}{\Afootnote{\emph{sic}}}.  \edtext{Ceit\emph{r}i}{\Afootnote{\emph{followed by \emph{meich} expunged; supply \emph{fogla}}}} lain .i. feiss l\'{a}ei, feis aidhche, ruireach aidhche, ⁊ airlim aidche cona t\'{a}rachtai\emph{n} iar maitti\emph{n}; lethfiach airlimi in laei a tairrsge in l\'{a}i, is\'{i} in fog\emph{ail} ceathr\emph{amth}an.  Tiugradh  ca\emph{ch} coimdi .i. in lai\emph{n}gfitir.  Cach coimrithe .i. in rith roin\emph{n}e.  Catch comhimridh .i. foimri\emph{m} iar n.x.m\emph{aid}, ɫ righe tar t\emph{r}achta co fis a et\emph{racht} imab-- cen taisealb\emph{ud}.  Scor cen taisilb\emph{ud}, is aithgi\emph{n} in\emph{n}.  in taisealb necnairc diama chin\emph{n}te co fogebhtha \'{e} fo cetoir imuich, \edtext{isl\emph{an}}{\Afootnote{is slan}}.  Madh qun\emph{n}tob\emph{ar}t, is leathaithg\emph{in}; diama chin\emph{n}te na foghebtha, is aithghin comhlan.  Asreanar tig\emph{r}adhus .i. is ar daighin marbhtha rochuir iatt. ɫ ecell .i. gnimh borblachais do imirt fair iarna thoirmescc dhe. Torbeach do \edtext{.r.}{\Afootnote{rala?}} .i. nad iar tairbiche an diumusa do \emph{m}uin\emph{ntir} fl\emph{ath}a no ecl\emph{as}a. 
        \pend
  \endnumbering
  \end{Leftside}

 \begin{Rightside}
    \beginnumbering
    \pstart  
    how should/shall I judge animal trespass .i. a follower's lamentation.  let you take if [\ldots] .i. distrait bruigh-law an oral contract which is a proper legal arrangement .i. according to doing his crime.  in return after legal warning .i. a crime is after legal warming [\ldots] of a sack or of the trespass by person.  four full penalties .i. animal trespass of day, trespass of night, rapid by night, night jumping forward in cattle trespass so that [it] is not found after morning;  half fine [for] trespass by cattle of day which animal trespass of the day, it is quarter penalty [for] the trespass by animal.  last legal person responsible of hobbled cattle, .i. the fetter between fore and hind leg, competing .i. the counting of a part. Every [\ldots] .i. act of using without authorization after [\ldots]. A kingship over strength with knowledge of his weakness [\ldots] without [\ldots].  Unharnessing without [\ldots], it is restitution in it.  The act of assigning [\ldots] certain so that you should find it immediately outside, which is exempt.  if doubt, is it s half-restitution; [\ldots] certain you should not find, which is equally full restitution.  last legal person responsible be paid compensation .i. it is because you may kill [that] he had sent for them.  or an unforeseen peril .i. a deed of harshness to use on it after (on account of) hindering from him. profitable worker [\ldots] .i. it is not after [\dots] the arrogance for a household of a lord or church.
    \pend

   \endnumbering
  \end{Rightside}
  \Pages

  \begin{Leftside}
    \beginnumbering
    \pstart
    \textbf{T}igradhus ca\emph{ch} sell\emph{aig}(?) .i. Ma dethb\emph{ir}us amuigh ⁊ thall \edtext{isl\emph{an}}{\Afootnote{= is slan}}. Mad ind-.x.b\emph{ir}us in\emph{n}tib a ndis, is fiach seillcechta.i. t\emph{r}ian aithgina, ɫ aithgin, ɫ ceathraime diri re taebh aithgina.  Mad .x.b\emph{ir}us in dara de ⁊ indeithb\emph{ir}us in ar.ii., iss i\emph{n}an\emph{n} ⁊ da in.x.b\emph{ir}us. Madh .x. b\emph{ir}us imuigh \edtext{im\emph{morro}}{\Afootnote{i\={m} = im\emph{morro}}} ⁊ con\emph{n}tob\emph{air}t tall, is leth ca\emph{ch}a rain\emph{n}i seillcechta. Co b\emph{er} \edtext{\emph{breith}}{\Afootnote{.b.}} im \edtext{\emph{techtae}}{\Afootnote{.t.}} mbrogha? .i. tir \edtext{\emph{dibaid}}{\Afootnote{.d.}} | coitcheann so, ⁊ is edh dleaghar dibh uile a ime etorr\emph{u},⁊ neach dibh iar si\emph{n} timcheall dia forrumadh; ⁊ \edtext{isl\emph{an}}{\Afootnote{= is slan}} d\'{o} cia foghluighid ceathra and itir da uair na forraimhe; ⁊ muna derna it\emph{ir} an forrumhe in la-sin, is fiach duinechaithe cunnthab\emph{air}tghe fair ind cen \.{f}is na ni\emph{nn}ili dorighne in fog\emph{gail}; ⁊ dia fi\emph{nn}ta iatt, icaitt fia\emph{c}h fo aic\emph{ned} in inait asa tangata\emph{r} do dhe\emph{n}amh na fog\emph{la}. ⁊ isl\emph{an} eisimh, ar is i\emph{n}bleag\emph{an}; ɫ dona co mbeith smacht air-si\emph{m} \={s} a saera ar aithg\emph{in} fe\'{o}ir nama. \edtext{\emph{no}can}{\Afootnote{ɫcan = \emph{no}can}} orga\emph{r} .i. in .iiii.aird no culaird. ɫ cidh nodonoirg .i. cia ceithir dona .x. ceithribh. ut est geilit eich co huir, claidhit muca co g\emph{r}ian, ⁊ beirit beich blasa ⁊ tora eisi. 
    \pend

    \pstart
    \textbf{C}o beir \edtext{\emph{breith}}{\Afootnote{.b.}} im \edtext{\emph{techtae}}{\Afootnote{.t.}} cominghairi.  in comi\emph{n}ghairi, dligh\emph{id}h ca\emph{c}h dia dh\'{e}namh ar lin ceathra, ⁊ na roibh d\emph{r}ochmil i nag\emph{aid} deighmhil in\emph{n}te, ar is olc d'fer in d\emph{r}ochmil deighaithgi\emph{n} do ic f\emph{r}i neach ⁊ na hicth\emph{ar} ris \={s} d\emph{r}ochaithg\emph{in}; ⁊ ni theit neach a dlighi corusa fi\emph{n}e cengo dern\emph{n}aidh an chominghairi mu\emph{n}ab co\emph{m}aithi na ceathra; ⁊ ar aidhe madh \'{e} fer in d\emph{r}ochmil rig a les a dhenomh ris, aslaighi dlighi a dheano\emph{m} ris, ar is do fein a olc.
    \pend
    \endnumbering
  \end{Leftside}

  \begin{Rightside}
    \beginnumbering
   \pstart
   A last legal person [is] each a witness .i. if [\ldots] is out and yonder which is exempt from liability.  If [\ldots] is in them their two people, it is an obligation of witnessing .i. a third part restitution, or restitution, or fourth part of honor price with side-restitution.  If it is [\ldots{}] in second of two and [\ldots{}], it is equal and two [\ldots{}]. if [\ldots{}] is outside however and a doubt then, it is half of each witnessing part.  how should/shall I judge concerning legal rightness of land? .i. land which is free from liability | owned in common, and it is owed to all [of] them out of a fence/hedge between them, anyone to them after that boundary from their cattle-guard; and it is free from liability for him though he plunders four in it between two hours of stalking; if animal paw between the stalking enters into a compact, it is a doubt on him the debt of a harsh man in it without knowledge [\ldots] which he has done the trespass; and when they must know, and the inherent quality of the debt [is] made [in] restitution out of which they came to do [\ldots{}]. and a brace is free from liability, since it is distrait of a surety, or for it is so that being a penalty on him but which is not free on restitution of merely grass.  he is not slain .i. in all directions or the four nearest townlands. or although [\ldots{}] .i. what beast for not ten fours. as a horse grazes until, pigs dig with a sun, a sown of bee(?) taste(?), and assistance(?) out of it. for, it is evil for a man of the evil beast good restitution which pays against anyone and he is not paying against it bad restitution; 
    \pend
  
    \pstart
    how shall/should I judge concerning legal grazing. grazing, each entitlement from his doing on a full four(?). and an evil beast was not in front [of] a good beast therein, you do not go [with] anyone his entitlement of legal regulation of a corporate kin group without that [\ldots{}] the grazing if it is not equally good of the four; and on night if a man the evilbeast of a king his redress which is doing against him, an influence of entitlement which doing a king, for it is for yourself his evil. 
    \pend
    \endnumbering
  \end{Rightside}
  \Pages

  \begin{Leftside}
    \beginnumbering
    \pstart
    \textbf{C}ia doneithear dono faill cometa in\emph{n}ti, isl\emph{an} fo aicn\emph{ed} inti dorighne, cid urr\emph{ad} cia deor\emph{ad} cidh murchurtha, ⁊ ni lan cin\emph{n}tech airithe.  Sl\emph{an} do cidhbe aidhe nomber \={s} cu \emph{r}oairidhi fo cetoir, ⁊ mu\emph{n}a airidhe is fiach faill\emph{e} di\emph{n} \edtext{beommil}{\Afootnote{beo in mil \emph{O'C., a possible reading ; if so, ommit \emph{in}.}}}.
    \pend

    \pstart
    \textbf{M}ad imin mart dogn\'{e} in faill, is fiach faille do\emph{n} mhart an\emph{n}-sidhe.
    \pend
    
    \pstart
    \textbf{I} Cuairt cominghairi atait si\emph{n} uili; ⁊ mad r\emph{uc} iad a .uii.archuairt comi\emph{n}gairi, ⁊ co\emph{m}aith an f\'{e}r thall ⁊ \edtext{amuich}{\Afootnote{\emph{separation of this from the following par.}}}.
    \pend

    \pstart
    Cach n\'{i} isl\emph{an} do thall isl\emph{an} amuich; ⁊ ciamadh aith\emph{gin} tall, ni teid t\emph{ar} aithg\emph{in} bheos imich.  Ma mesa in fer amuich oldas tall, is fia\emph{ch} a \.{f}atha fair-siu\emph{m} ina bremph{eith} amach.  Mad fearr in fer amuich, ca\emph{ch} n\'{i} \edtext{asl\emph{an}}{\Afootnote{= as slan}} d\'{o} than isl\emph{an} amuich. ɫ ar thairghill\emph{e} .i. geall t\'{o}irithnech d\'{o} do b\emph{re}t(?) d\'{o} t\emph{ar}a cheann co reithean\emph{n} fuill\emph{em} ngill d\'{o} ris, ⁊ ni fearr leis fuillem .ii. dobera na logh d'inga\emph{ir}e do de\emph{n}am dh\'{o}, ⁊ ni dligi asl\emph{aig}his air a \edtext{deanomh}{\Afootnote{\emph{followed by scribal note.}}}.
    \pend
    \endnumbering
  \end{Leftside}

  \begin{Rightside}
    \beginnumbering
   \pstart
   What [\ldots] moreover neglect of guarding in it, which is free from liability under inherent quality he who [\ldots] who is a native, although a foreigner who is come over seas, and full fixed is not certain.  Safe from liability for however it is a special repository for a surety [\ldots] but with keen guards immediately, and if a house hold item is payment of neglect out of the living animal(?). 
     \pend

     \pstart
     if it is a vow which should do the negligence, it is a payment of negligence for the ox meat in it.
     \pend

     \pstart
      in a circuit of grazing which is all that; and if it is that which had bore them of 7 circuits of grazing, and that the grass of the defendant is equally good.
     \pend

     \pstart
      Each which is not free from liability for the defendant is free from the plaintiff; what if it is restitution [of/by?] defendant, he does not go across restitution against a plaintiff. if judging the defendant than a plaintiff, payment is his reason on it in his judgement out. if the defendant is best, every which is not out of liability to a plaintiff which is exempt out. or on a first surety .i. a revealing surety to him [\ldots] to him in return for  [\ldots] profit of a surety for him against it, and profit is not best with him .ii. the compensation should/shall carry for herding to do to him, and it is no entitlement of persuading on him to do. 
     \pend
    \endnumbering
  \end{Rightside}
  \Pages

\begin{Leftside}
    \beginnumbering
    \pstart
    | \textbf{C}o ber \edtext{\emph{breith}}{\Afootnote{.b.}} im rathosa \edtext{\emph{ruip/rop}}{\Afootnote{.r.}}? ni fuil ni o rop .c.ci\emph{n}tach ina .c.fog\emph{ail} muna rai\emph{n}ic co hothrus ɫ aithg\emph{in}, \={s} asl\emph{an}; ma minchneadh .ii. fearas ina dhiaidh, teitt in c\'{e}dfog\emph{ail} a naiream cinaidh, ⁊ rop decintach inosa \'{e} ⁊ biaidh leithfiach isi\emph{n} minchneidh tanasi cengo rainic othrus no aith\emph{gin}.
    \pend

    \pstart
    \textbf{A}r is comfiachach in braithemh ⁊ in gonaidh.  Ceathror comfia\emph{c}hach .i. airi echto ⁊ fear b\emph{r}aith ⁊ sell\emph{ach} la\emph{n}fiach ⁊ fer laimhi; \={s} mad cuibhdhi iad, com\'{i}caitt uili .ui. cum\emph{al}a diri, ⁊ icaidh fer laimhe a oenur in cu\emph{mail} aithg\emph{en}a. 
    \pend

    \pstart
    Mad ecuibhdi \edtext{im\emph{morro}}{\Afootnote{i\={m}}}, icaidh cach ae dib \edtext{.iii.}{\Afootnote{.ui. \emph{O'C, but more like} .iii.}} cum\emph{al}a ⁊ ica fear laime a oenur .uii. cum\emph{al}a; ⁊ muna tarrus fear laimhe, is e fear b\emph{r}aich icos aithg\emph{in}. \edtext{ut dixit}{\Afootnote{ut d\={x}}}. mecon cina cubusb\emph{r}ath.  ⁊ idbeir cain: int\'{i} nodmairnd is \'{e} adrean diri ⁊ aithg\emph{ein} mad anurr\emph{aid} mairndi. A rofitir cein ro\={b} .i. fear b\emph{r}aith. Aritruisith\emph{er} .i. sell\emph{ach} lain. frisitroraidh .i. sell\emph{ach} leithe. conimrol\emph{aid}(?) \edtext{.i.}{\emph{supply \emph{sellach}?}} cethraimhe, ɫ fear caimthechta.
    \pend

    \pstart
    \textbf{C}o b\emph{er} br\emph{eith} im ai\emph{n}cesa .i. im in\emph{n}ceas br\emph{eith}e. Nosb\emph{er}a .i. f\emph{or} do menmain. iar fochmarc mis .i. mi f\emph{r}i fochmarc ceithri leithe co fis cacha leithi .i. .uii.muin ca\emph{ch}a leithi.
    \pend

    \pstart
    \textbf{C}o b\emph{er} \edtext{\emph{breith}}{\Afootnote{.b.}} iter saer .i. g\emph{r}adh fl\emph{ath}o. ⁊ doer .i. g\emph{r}adh \.{f}ei\emph{n}e.  conairlithear f\emph{r}i saer .i. airail fair deitti\emph{n} dlighi don doer. ni mhairnte nach doer .i. i mbreith. 
    \pend
    \endnumbering
  \end{Leftside}

\begin{Rightside}
    \beginnumbering
   \pstart
   how shall/should I judge concerning rent surety? There is no blood, not from it should be a guilty part in his attack if not he attains with sick-maintenance or restitution, but out of liability; if (it is) a small wound .ii. conduct following the hundred attacks goes their sum of guilt, and it should be a double crime [\ldots] him and half-payment will be in the small would [case?] with that he arrived [at] sick-maintenance or restitution. 
   \pend

   \pstart
    For the betrayer and killer are(?) equally liable. with the four of them equally liable .i. man of betrayal and full penalty witness and man of power(?); but if they are proper, all neighbors six female slaves of honor price, and the man of power pays himself the female slave of restitution. 
   \pend
   
   \pstart
    If it is disharmony moreover, every fault pays to them 3 female slaves and a man of power makes compensation himself 7 female slaves; and if no bond imposed on a man of power, it is a man of betray who pays restitution. as he said. a root(?) complicit through previous knowledge. the aforementioned who betrays and it is [\ldots] of honor-price and of restitution if he is an outlaw of (by?) betrayal. When he knew enduring(?) [\ldots] .i. man of betrayal. [\ldots] .i. full witness. [\ldots] .i. half witness. [\ldots] .i. fourth part of penalties, or a man of crooked-going.  
   \pend

   \pstart
   how should/shall I judge concerning crime .i. concerning crime of a judge.  he will bear it .i. on your attention. after inquiring for a month. a month against inquiry of four halves without knowledge of every half .i. 7 treasure(?) of every half.  
   \pend

   \pstart
   how shall/should I judge between freeman .i. grade of lordship and unfreeman .i. grade of F\'{e}ne. he takes council against a freeman .i. before disgrace(?) upon him recognizing of entitlement for the unfreeman. it is not of betraying that is not an unfree man .i. in judging.
   \pend
    \endnumbering
  \end{Rightside}
  \Pages

\begin{Leftside}
    \beginnumbering
    \pstart
    \textbf{C}o b\emph{er} breith i\emph{m} \edtext{\emph{techtae}}{\Afootnote{i.e. techtae (\emph{or} tuitim ?)}} ngill .i. ca\emph{ch} inadh a fuil tuitim in ghill isna fia\emph{c}haib, ma ta treba\emph{ir}i re aisic na himforc\emph{r}adha, is a haisic, ⁊ muna fuil, is dilsi na himarc\emph{r}adho. Cach inadh \edtext{\emph{immorro}}{\Afootnote{i\={m}}} a fuilit feich tar eisi uili, is dilsi na hi\emph{m}arc\emph{r}adha, ce beith treba\emph{ir}e cenco b\'{e}[\hspace{2mm}]am\emph{ail} is dilis in colann dia lean an i\emph{m}arc\emph{r}aidh \edtext{ce beith t\emph{r}eb\emph{air}e cenco b\'{e}}{\Afootnote{\emph{perhaps dittog., more especally since followed by \emph{a\={m} is dilis in colonn dia lean an imf\emph{or}c\emph{r}aidh} cancelled by underlining}}}, ⁊ feich dara eisi. 
    \pend

    \pstart
    \textbf{L}andire do raith .i. boairi medonach .i. lanenecl\emph{ann} an boaire medonaidh is \'{e} leithenecl\emph{ann} an airech d\'{e}sa. leithdiri do aitiri .i. aire d\'{e}sa. ni teid aidire \={s} co cr\'{o} .i. co c\emph{r}o a aei ɫ co c\emph{r}o in f\emph{ir} mairbh .i. is f\emph{r}i hannscuichi b\'{i}te na haitire; toibhghid na comorb\emph{ai} no \'{i}cait. Teid rath f\emph{or} coma\emph{r}b\emph{u} co \edtext{\emph{nomadh}}{\Afootnote{ɫmhadh}} n\'{o} .i. co neimbeith nae, ɫ coruici .x. huadh.
    \pend
    
    \pstart
    IS aire theidi for comorb\emph{u} co fo do sond, uair bi f\'{u}aidri forsan tir sonn, cidh cor da sochonn \edtext{ɫ}{\Afootnote{\emph{breaks off incomplete owing to chasm in MS}.}} 
    \pend
    \endnumbering
  \end{Leftside}

\begin{Rightside}
    \beginnumbering
   \pstart
   how shall/should I judge concerning lawful pledge[s] .i. every place which is falling due of the pledge[s] in the obligations, if it is a security against [\ldots] which is his [\ldots], and if it is not blood, which is a right over the [\ldots]. every place moreover which his [\ldots] of obligation on behalf of all, which is a right over [\ldots], although judging security with that should be[\hspace{2mm}]should be like which is belonging to the pledged object if he prosecutes a claim arising out of [\ldots] and of obligation after it.  a surety goes on heirs to the ninth generation .i. with poisoned-being human, or 10 disgrace from it.
    \pend

    \pstart
    full honor price for a surety .i. the intermediate \emph{boaire} .i. half honor price of the \emph{airech d\'{e}sa}. half honor price for a surety .i. \emph{aire d\'{e}sa}. the surety does not go but with hereditary property .i. with hereditary property of his lawsuit or with hereditary property of the dead man .i. it is against immovables which are the sureties. the side pledge of the heirs who make compensation. 
    \pend

    \pstart
    It is a aire on heirs [\ldots] a stake, an hour is an impunging a contract on the land here, although an oral contract for a person of sound intelligence or \ldots
    \pend
    \endnumbering
  \end{Rightside}
  \Pages
\end{pages}

\section{Notes}

\begin{enumerate}
  \item[sruith] i-stem masc. \enquote{elder, venerable}
  \item[Co] gem. conj. part. \enquote{how} GOI p. 290
  \item[ber] 1st pers. pres. subj. conj. or 1st person fut. indict. act. (no length mark)
  \item[cionnus] from cindas (d) \enquote{how?}
  \item[berat] 3rd per. pres./fut. subj. conj. \enquote{should/shall bear}
  \item[b\emph{r}eatha] \={a}-stem f. acc. pl. \enquote{judgements}
  \item[b\emph{er}o] 1st pers. subj./fut. act. \enquote{I shall/should bear} 
  \item[nip] 3rd pers. pres. subj. neg. of is \enquote{so that it may not be}
  \item[sl\emph{an}] DIL S 260 \enquote{exempt, non-liable, safe}
  \item[saithi] \enquote{throng, multitude}
  \item[diub\emph{air}t] acc. sing. \={a}-stem frm \enquote{fraud} -- d\'{i}upart
  \item[mboeth] from b\'{a}eth (also b\'{o}eth) \enquote{legally incompetent}
  \item[hiadhat] \'{i}adaid \={a} 3rd sing. pres. indic. act. conj. trans \enquote{closes, fastens, clasps, joins} intrans. \enquote{closes (a transaction)}
  \item[coruice] co + ruccae io, masc. \enquote{shame, disgrace}
  \item[dofuasl\emph{aic}e] from do-f\'{u}asailci 3rd perl pl. \enquote{sets free, dissolves}
  \item[doch\emph{ur}] dochor \enquote{disadvantageous contract}
  \item[memar] \enquote{member of a community under a head or chief}
  \item[saire] from sa\'{i}re, i\={a}-stem f. \enquote{privileges, immunity, exemption}
  \item[resin] r\'{e}sin? GOI \S 851 (pg. 527) \enquote{before the}
  \item[ini] in-n\'{i}(?) GOI pg. 291
  \item[rannait] pres. 3rd pl. of rannaid \enquote{divides}
  \item[feri\emph{ann}] o-stem, neut. or masc. \enquote{land, domain, territory}
  \item[ndilsi] (d\'{i}lse) \enquote{security, pledge, warrent}
  \item[haititiu] \enquote{acknowledging} (esp. property rights)
  \item[annfesa] \enquote{ignorance} see also \enquote{aititius ainfesa in codnaid} v496.7
  \item[codnach] \enquote{sensible adult}
  \item[do choinn] \enquote{in return for}
  \item[essl\emph{ain}] (eass + lain) l\'{a}n ess/ass pre. \enquote{out of [the] whole}
  \item[leithchil] cil \enquote{fault} [\={a}-stem, fem.?] \enquote{wrong judgement}
  \item[cloen] gen. sing of cl\'{a}en \enquote{unjust, iniquitous}
  \item[em-teclamtha\emph{r}] teclaimid(?) \enquote{collects, gathers, brings together} Mid. Ir.
  \item[cutuma] cutrummae \enquote{equal}
  \item[Arnacon] final conjunction neg. GOI \S 898 \enquote{that not}
  \item[dearba\emph{r}a] from do-opir see DIL D2 352.011 subj. pres. 3rd sing. \enquote{defraud} 
  \item[ailaile] pron. \enquote{another, the other}
  \item[amnech] am + nech \enquote{indeed anyone} 
  \item[aml\emph{aid}-seo] \enquote{Thus, so, accordingly} Mid. Ir.
  \item[treb\emph{air}e] i\={a}-stem fem. \enquote{security, guaruntee}
  \item[easbuidh] esbaid, esbud \enquote{lack, loss, deficency, defect}
  \item[tessb\emph{aid}] vn. of do-esta \enquote{being lacking, missing, absent}
  \item[cor] vn. of fo-ceird \enquote{putting} legal meaning \enquote{verbal contract} or later \enquote{surety, guarantor}
  \item[dhiss] from d\'{i}as \enquote{two, dual}
  \item[echtrann] \enquote{foreign, stranger}
  \item[m\emph{ain}e] gen. pl. of ma\'{i}n \enquote{cattle, stock}
  \item[comlinat] vn. \enquote{act of filling up, completing}
  \item[fol\emph{ad}] (folud) in Laws \enquote{means, materials} or otherwise by which a person sustains his functions or liabilities
  \item[loighi] (lige) \enquote{lying down} with adj. has many meanings
  \item[co\emph{m}ann] \enquote{cattle-pound}
  \item[innrigthigh] DIL R 69.85 \enquote{forearms?} or in righ tigh \enquote{the king's house}
  \item[ath\emph{gabail}] \enquote{distraint}
  \item[cae] \enquote{visiting}
  \item[mann] \enquote{ingot} or \enquote{god-sent, manna, blessing} (doesn't make sense)
  \item[trosca] \enquote{fasting}
  \item[saer] \enquote{free-client}
  \item[nadma] \enquote{surety}
  \item[is saml\emph{aid}-seo] \enquote{thus it is}
  \item[cuiti] \enquote{some, a certain amount}
  \item[d'fas\emph{tad}(?)] do + fastad (see astud, fast\'{o}d, fostad) astud \enquote{act of holding back, detaining, fixing, establishing}
  \item[dociallath\emph{ar}] meaning doubful \enquote{forgets, neglects, impairs}
  \item[dogne] (do-gn\'{e}) 2nd/3rd pers. pres. sub. \emph{which may do it}
  \item[adbul] mistake intensive prefix DIL A37.004
  \item[do reir] see r\'{i}ar -- \enquote{at the will of \ldots}, \enquote{on the authority of\ldots}, \enquote{according to\ldots}
  \item[firine] from f\'{i}r\'{e}n gen. sing. \enquote{just, righteous}
  \item[Ar is] \enquote{for} conj. see GOI \S 906 \enquote{for, \ldots}
  \item[firsiu] \enquote{strength} (archaic/poetic)
  \item[f\emph{er}tsechus] fert + sechus = righteous?
  \item[n\emph{ir}t] gen. of nert \enquote{strength}
  \item[fia] from f\'{i}ad \emph{in the presence of}
  \item[comtoiridhne] com + tairiden \enquote{leading (water)} or \enquote{mill-race} so \emph{common water}?
  \item[ast\emph{ad}] (astaid) see ad-suidi vn. \enquote{legally fixes, makes permanent, binds, establishes}
  \item[\emph{f\emph{or} fonaidm}] fo-naidm vn. of fo-naisc DIL F 287.061 \enquote{pledge, bond guarantee} 
  \item[ar aird] DIL A 182.45 \enquote{at it}
  \item[degnadmaim] deg-nadmaim from d\'{e}g \enquote{because of, in return for} + nadmaim dat. sing. neut \enquote{act of binding, a bond, pledge}
  \item[F\emph{or}congair] 3rd sing \enquote{commands} subordinate clause of thing commanded
  \item[aurgair] imper 2nd sing DIL A 397.011 \enquote{forbid}
  \item[f\emph{or}cong\emph{r}a] io-stem masc. \enquote{commands, summons}
  \item[man] mad(?) \enquote{good, fortunate}
  \item[rourgair] ro + urgair 3rd pers. sing. past (perf.?) of ar-gair \enquote{forbids, prevents}
  \item[in cuir] gen sing of cor legal meaning \enquote{contract}
  \item[leicit] 3rd pl. pres. indic. activ of lecid \enquote{leave}
  \item[taith\emph{mech}] vn of do-aithbig \enquote{dissolving, annulling a contract}
  \item[Eirne] 3rd sing. of ernaid \enquote{bestows, grants}
  \item[contuaisi] indict. pres 3rd sing. of com-t\'{u}aisi \enquote{listens, hears}
  \item[nurrtuasacht] from urrtuasacht \enquote{keeping silent, not revealing}
  \item[Fuidhri] DIL F 467.050 \enquote{tenant}
  \item[cundu] \enquote{friendship}
  \item[indile] \enquote{property}
  \item[dethb\emph{ir}us] deth \enquote{good} + birus (?)
  \item[miss] acc. sing. of m\'{i} \enquote{month}
  \item[sacar\emph{ad}] vn. of scaraid \enquote{act of sperating}
  \item[Mad] pres./past. subj. \enquote{if it were}
  \item[indethb\emph{ir}us] \enquote{not good}?
  \item[a ndis] \enquote{their standing}?
  \item[\.{f}e\'{o}ir] gen. sing. of f\'{e}r \enquote{grass} obj of vn.
  \item[tuco] ?
  \item[infasas] from in\'{a}said \enquote{grows, increases} (inf\'{a}sas?)
  \item[foraind] 1st pl. for + pron. suffix \enquote{on us}
  \item[barr] \enquote{end}
  \item[doratt] 3rd. sing. rel. perf. act. \enquote{has given}
  \item[iasacht] \'{i}asacht \enquote{a loan, lending}
  \item[arathir] ?
  \item[techta] t\'{e}chtae \enquote{legal rightness}
  \item[dib\emph{aid}] gen. sing. \enquote{inherentince}
  \item[Comairme] \enquote{proportional stock, equal stock}
  \item[cutrumas] co + trummae?
  \item[airme] from \'{a}irem gen. sing. and nom. pl. \enquote{counting}
  \item[Comecor] ?
  \item[Comairimh] from com\'{a}irem \enquote{proportional stock
  \item[ruc] pres. subj. of beirid see GOI \S 902}
  \item[ima\emph{r}craidh] from immarcraid \emph{excess, surplus}
  \item[mil] m\'{i}l \enquote{animal}
  \item[ecuttruma] ecut + truma (heavy?)
  \item[a\'{i}s] gen. sing. of \'{a}es \enquote{age, years}
  \item[nag\emph{aid}] n-agaid \enquote{face/in front of}
  \item[ainmil] ain + mil \enquote{driving animal}
  \item[fiach] payment due, obligation
  \item[duninicaiche] DIL C 88.003 \enquote{of tresspass by person}
  \item[tairrsce] \enquote{animal tresspass}
  \item[caei] \enquote{act of lamentation}
  \item[tairis] tairise \enquote{of a follower}
  \item[Gab\emph{ae}(?)] 2nd pers. sing. pres. subj./fut. of gaibid
  \item['troiris] ?
  \item[bruighrechta] DIL I 265.070 \enquote{bruigh-laws}
  \item[chinaidh] \enquote{fault, crime}
  \item[foc\emph{ru}] frmo focra(e) \enquote{a warning of legal action}
  \item[ataitt] ? at-t\'{a}
  \item[meich] gen. sing. of m\'{i}ach \enquote{sack}
  \item[Ceit\emph{r}i] \enquote{four}
  \item[lain] legal sense \enquote{full penalties/amount}
  \item[feiss] \enquote{tresspass}
  \item[ruireach] \enquote{king, chief} or \enquote{swift, rapid}
  \item[airlim] vn of ar-ling \enquote{act of jumping forward in cattle tresspass} DIL A 225.074
  \item[cona] \enquote{so that not} GOI \S 896
  \item[t\'{a}rachtai\emph{n}] \enquote{finding of a guilty animal}
  \item[is\'{i}] is + \'{i}
  \item[ceathr\emph{amth}an] from cethramthu \enquote{quarter penalty}
  \item[Tiugradh] \enquote{last legal person reponsible}
  \item[lai\emph{n}gfitir] \enquote{the fetter between fore and hind legs}
  \item[coimrithe] from Comrith \enquote{competing} see DIL C 410.032
  \item[in rith roin\emph{n}e] \enquote{the counting (liabilities?) of a part}
  \item[foimri\emph{m}] \enquote{act of using without authorization}
  \item[t\emph{r}achta] \enquote{strength}
  \item[co fis] \enquote{With knowledge} DIL F 152.033
  \item[et\emph{racht}] \'{e}tracht \enquote{lack of strength}?
  \item[Scor] \enquote{act of unyoking, unharnessing} or \enquote{paddock}
  \item[taisealb] from taiselbad \enquote{act of assigning} perhaps \enquote{stating to be the property of another}
  \item[necnairc diama] ?
  \item[fogebhtha] from fo-gaib/geib sec. fut. 2nd sing \enquote{you should find/discover}
  \item[imuich] \enquote{outside} see immaig
  \item[qun\emph{n}tob\emph{ar}t] cuntabart (thanks to Simon Rodway) \enquote{doubt}
  \item[comhlan] \enquote{equally full}
  \item[Asreanar] from as-ren (asrenar) pass. sing. \enquote{pays away fine or compensation}
  \item[ar daighin] \enquote{because of} or \enquote{on account of}
  \item[marbhtha] 2nd pers. sing. imperf. ind./past. sub.
  \item[rochuir] leniting relative? ro-cuir? or \enquote{sent for} see cuirid DIL C598.062
  \item[ecell] \'{e}cell \enquote{unforeseen peril, trap}
  \item[borblachais] \enquote{fiercness, harshness, violence}
  \item[do imirt] \enquote{act of using, employing}
  \item[thoirmescc] tairmesc \enquote{hindering, restraining, preventing prohibiting}
  \item[Torbeach] \enquote{profitable worker}
  \item[tairbiche] ? 
  \item[diumusa] see d\'{i}ummus \enquote{vainglory, arrogance, pride}
  \item[sell\emph{aig}(?)] \enquote{witness} acc. sing.
  \item[thall] \enquote{yonder}
  \item[in\emph{n}tib] 3rd pl. dat. + prep. i \enquote{in them}
  \item[ndis] \enquote{a pair of people}
  \item[seillcechta] \enquote{witnessing}
  \item[t\emph{r}ian] \enquote{third part}
  \item[ceathraime] from cethramthu \enquote{fourt part}
  \item[diri] io-stem neut. \enquote{honor price}
  \item[taebh] \enquote{side} or \enquote{as well as}
  \item[dara] \enquote{the second of two}
  \item[i\emph{n}an\emph{n}] from inunn \enquote{alike, equal}
  \item[rain\emph{n}i] gen. sing. (rainne) of rann \enquote{part}
  \item[mbrogha] gen. sing. of bruig \enquote{land}
  \item[coitcheann] \enquote{public, owned in common} 
  \item[dleaghar] \enquote{it is due to them} see DIL D2 160.74 
  \item[a ime] \enquote{out of a fence/hedge}
  \item[timcheall] \enquote{boundary}
  \item[forrumadh] from forruma \enquote{a guard on cattle} 
  \item[foghluighid] (fogluigid) from Dineen fo\.{g}lu\.{g}im - \enquote{I plunder} (see DIL F 228.066 Fogal) 3rd sing \enquote{he plunders}? 
  \item[forraimhe] from foram \enquote{moving towards, stalking, hunting}
  \item[muna] Mod. Ir. m\'{a} + neg. \enquote{if not}
  \item[derna] \enquote{animal paw}? DIL D2 40.072
  \item[in la-sin] \enquote{enters into a compact} DIL I 271.055
  \item[duinechaithe] \enquote{man} + \enquote{hardness, stifness}
  \item[cunnthab\emph{air}tghe] \enquote{doubt} but what is -ghe? 
  \item[dorighne] 3rd pers sing. perf. act. of do-gn\'{i} (do-rigni $\rightarrow$ dorighne)
  \item[fi\emph{nn}ta] late verbal of necessity of ro-finnadar or ro-fitir (see DIL F 145.046) \enquote{must know}
  \item[icaitt] icad for \'{i}ccaid \enquote{make resitution}
  \item[fo aic\emph{ned}] \enquote{inherent quality}
  \item[inait] gen. sing. \enquote{of place,spot} maybe \enquote{indemnification}
  \item[eisimh] from essem \enquote{brace or strap use to fasten one yoke to}
  \item[i\emph{n}bleag\emph{an}] inbleogan vn of in-omblig \enquote{distratin of a surety}
  \item[smacht] \enquote{penalty, fine}
  \item[nama] \enquote{only merely} or \enquote{n\'{i} \ldots nama} $\rightarrow$ \enquote{not only \ldots but} 
  \item[\emph{no}can] emphatic \enquote{no}
  \item[orga\emph{r}] from oirgid pass pres. \enquote{is slayed}
  \item[.iiii.aird] cetharaid \enquote{in all directions}
  \item[culaird] from c\'{u}laird \enquote{the four townlands nearest the cetharaid}
  \item[nodonoirg] no-do-n-oirg?
  \item[ceithir] see cethir \enquote{four-footed animal/beast}
  \item[geilit] 3rd sing. pres. indic. act. (gelid) \enquote{grazes, consumes}
  \item[claidhit] (claidid) \enquote{digs, excavaates} 3rd pl.
  \item[beirit] see birit DIL B 105.036
  \item[tora] gen. sing. ? of t\'{o}ir \enquote{helping, persuers}
  \item[cominghairi] from cominghaire \enquote{grazing}
  \item[lin] \enquote{full number} (maybe \enquote{legal party}?)
  \item[d\emph{r}ochmil] droch \enquote{evil, bad} + mil (m\'{i}) \enquote{beast}
  \item[deighmhil] dag \enquote{good} + mil (m\'{i}l) \enquote{beast}
  \item[in\emph{n}te] \enquote{into it}/\enquote{therein}
  \item[hicth\emph{ar}] pass. conj. of \'{i}ccaid \enquote{paying}
  \item[theit] 2nd pl. conj. pres. indic. act. of t\'{e}it (misspelled -t\'{e}t?)
  \item[corusa] (c\'{o}rus) \enquote{regulation, arrangement} (legal)
  \item[cengo] from cenco \enquote{without that}
  \item[dern\emph{n}aidh] ?
  \item[mu\emph{n}ab] same as manib ma + neg. 3rd sing \enquote{if he is not}
  \item[co\emph{m}aithi] \enquote{equally good}
  \item[aidhe] \enquote{night}
  \item[a les] \enquote{his relief/redress}
  \item[aslaighi] (aslaige $\rightarrow$ aslach?) \enquote{persuading, influences, tempting}
  \item[doneithear] do-neithear?
  \item[faill] \enquote{neglect}
  \item[cometa] from comi{e}t \enquote{act of keeping, guarding, a guard}
  \item[dorighne] ?
  \item[urr\emph{ad}] from aurrae \enquote{person wh ois native to t\'{u}ath} gen. sing? 
  \item[deor\emph{ad}] \enquote{outsider} 
  \item[murchurtha] from murchuirthe (-choirthe) \enquote{foreigner, come over seas} 
  \item[cin\emph{n}tech] \enquote{fixed}
  \item[airithe] from \'{a}irithe \enquote{certain, sure}
  \item[do cidhbe] \enquote{for however it} DILL \'{e} (II) E 1.028
  \item[aidhe] from aite 3 \enquote{special repository for an object given as surety}
  \item[nomber] no-m-ber 1st subj./1st sing. fut. (?) rel. nas. cl. or infix pron.
  \item[\emph{r}oairidhi] \enquote{keenly guards} ro- of intensity
  \item[fo cetoir] \enquote{at once, immediately} 
  \item[airidhe] \enquote{household funishings} DIL A 212.022
  \item[imin] ad-gella 2nd sing. pret./perf. see DIL A 56.083 \enquote{vows, promises}
  \item[mart] \enquote{ox meat}
  \item[dogn\'{e}] from do-gn\'{i} 2nd sing/3rd sing pres. subj. \enquote{should do} rel.?
  \item[an\emph{n}-sidhe] ann \enquote{in it} + sidhe emphatic pron. DIL S 220.027
  \item[atait] 3rd pl. of at-t\'{a} \enquote{they are} rel.
  \item[amuich] \enquote{outside/plantiff} see DIL 117.054
  \item[bheos] see beus \enquote{still, further, moreover} 
  \item[mesa] vn of midithir \enquote{act of judging}
  \item[oldas] \enquote{than}
  \item[\.{f}atha] \enquote{cause, reason} gs. or np.
  \item[asl\emph{an}] \enquote{out of liability}
  \item[thairghill\emph{e}]  thair \enquote{before, in front of} + gill \enquote{surety} 
  \item[geall] \enquote{pledge, surety}
  \item[t\'{o}irithnech] \enquote{helpful, revealing}
  \item[do b\emph{re}t(?)] ?
  \item[t\emph{ar}a cheann] \enquote{on behalf of, in return for, because, notwithstanding, although}
  \item[reithean\emph{n}] ?
  \item[logh] \enquote{value, worth/pushiment, compensation}
  \item[inga\emph{ir}e] \enquote{hearding}
  \item[rathosa] (r\'{a}tho + sa?) \enquote{rampart} or \enquote{surety} gen. sing.
  \item[rop] 3rd sing. pres. subj.of is \enquote{it should be}
  \item[.c.ci\emph{n}tach] .c. ??
  \item[rai\emph{n}ic] 3rd sing. pret. of ro-icc \enquote{reaches, arrives, attains}
  \item[hothrus] \enquote{sick-maintenance}
  \item[minchneadh] min + cneadh (from cned) \enquote{fine, small} + \enquote{wound}
  \item[fearas] \enquote{conduct, activity, office}
  \item[naiream] n-aiream \enquote{counting, amount, sum}
  \item[decintach] \enquote{double crime}
  \item[inosa] ?
  \item[biaidh] fut. of at-t\'{a}
  \item[tanasi] ?
  \item[braithemh] \enquote{betrayer}
  \item[gonaidh] \enquote{pierces, wounds $rightarrow$ kills} vn?
  \item[Ceathror] dat. of accompanyment see GOI \S 251(3)
  \item[airi echto] ?
  \item[laimhi]  \enquote{hand/power}?
  \item[cuibhdhi] from cubaid \enquote{fitting, proper}
  \item[com\'{i}caitt] \enquote{neighbor}
  \item[ecuibhdi] from \'{e}ccuidbe \enquote{disharmony}
  \item[ae] \enquote{fault}?
  \item[tarrus] pret of do-arret \enquote{imposes a bond on}
  \item[icos] rel. of \'{i}ccaid
  \item[mecon] \enquote{a root}?
  \item[cina] \enquote{of crime}?
  \item[cubusb\emph{r}ath] cubus + brath \enquote{complicit through previous knowledge}
  \item[idbeir] 3rd singl indef. pron. 
  \item[cain] \enquote{law}
  \item[nodmairnd] pres 3rd sing. \enquote{betrays} DIL M 40.069
  \item[adrean] ?
  \item[anurr\emph{aid}] \enquote{outlaw}
  \item[mairndi] rel 3rd person(?) of mairnid
  \item[cein] \enquote{long, enduring, far, distant} 
  \item[Aritruisith\emph{er}] ?
  \item[frisitroraidh] ? 
  \item[conimrol\emph{aid}(?)] ?
  \item[cethraimhe] \enquote{fourth part of penalties}
  \item[caimthechta] \enquote{crooked-going} DIL C 62.057
  \item[Nosb\emph{er}a] class A infx 3rd sing. or 3rd pl. with no-bera 3rd sing. pres. subj./3rd sing. fut.
  \item[menmain] menma acc./dat. \enquote{thought, heed, attention}
  \item[fochmarc] \enquote{asking, inquiring}
  \item[muin] \enquote{gift treasure}
  \item[conairlithear] from con-airlethar \enquote{advises, takes council}
  \item[airail] air \enquote{before, for} + ail \enquote{foundation of law, gravestone, disgrace}
  \item[deitti\emph{n}] from d\'{e}itiu u-stem fem. acc.(?) \enquote{recognizing, granting}
  \item[mhairnte] gen sing \enquote{of betraying, informing}
  \item[tuitim]
  \item[aisic] epithet for a cauldron? DIL A 160.061
  \item[himforc\emph{r}adha] ?
  \item[dilsi] from d\'{i}lse \enquote{right over} DIL D2 110.064
  \item[himarc\emph{r}adho] see DIL T 177.028
  \item[inadh] \enquote{place}
  \item[fuilit] ?
  \item[b\'{e}] 3rd sing. pres. subj. of at-t\'{a}
  \item[dilis] from d\'{i}les \enquote{belonging to, own, appropriate}
  \item[in colann] from colainn \enquote{material subject of a bargain, principle, capital}
  \item[lean] DIL L 98.084 \enquote{prosecutes a claim arising out of}
  \item[medonach] \enquote{central, intermediate}
  \item[airech d\'{e}sa] lowest grade of nobility
  \item[co cr\'{o}] \enquote{with hereditary property}? or \enquote{gore}
  \item[aei] \enquote{law suit/litigation}
  \item[hannscuichi] \enquote{immovables}
  \item[b\'{i}te] rel. of at-t\'{a}
  \item[toibhghid] toibh \enquote{side} + gid \enquote{pledge}
  \item[comorb\emph{ai}] \enquote{heirs}
  \item[co \emph{nomadh} n\'{o}] \enquote{to the ninth generation}
  \item[co neimbeith] \enquote{with poisoned being}
  \item[nae] \enquote{human} DIL N 55.051
  \item[aire theidi] \enquote{freeman/noble} + teidi?
  \item[co fo do] ?
  \item[sond] \enquote{stake, post, beam, prop, missle, palisade}
  \item[f\'{u}aidri] f\'{u}adrad $\rightarrow$ f\'{u}aitriud \enquote{impugning a contract}
  \item[sonn] \enquote{here}
  \item[da sochonn] \enquote{for a person of sound intelligence} 
\end{enumerate}
\end{document}
